A Linear Solution of the Linearity Challenge

### 2025-12-25 paper/Congruence.lagda.tex
global revision

The datatype for $\pcong$ has one constructor for each of the structural
precongruence rules in \Cref{tab:semantics}. Since many aspects recur
repeatedly, we will focus on just a few representative rules starting from
\SComm.

@@

There are two non-trivial aspects worth commenting. The first one concerns the
proof $(\AgdaInductiveConstructor{>}~q)$ used in the
\AgdaInductiveConstructor{wait} process before structural pre-congruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{ch~⟨~>}~q~\AgdaInductiveConstructor{⟩}~P)$
    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
    it is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel (say $y$, of
    type $\Bot$) consumed by the \AgdaInductiveConstructor{wait} process.
    %
    We enforce the side condition $x \ne y$ of \SWait (which we left implicit in
    \Cref{tab:semantics}) imposing that the type $A$ in front of
    $A~\AgdaInductiveConstructor{∷}~\Context_1$ necessarily comes from the right
    partition of the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$
    through the use of \AgdaInductiveConstructor{>}.
\end{enumerate}

@@

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $(\AgdaInductiveConstructor{<}~r)$, meaning
that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
\AgdaInductiveConstructor{s-fork-l} except that
$(\AgdaInductiveConstructor{<}~r)$ is replaced by
$(\AgdaInductiveConstructor{>}~r)$.

>>>>

The datatype for $\pcong$ has one constructor for each structural pre-congruence
rule in \Cref{tab:semantics}. Since many aspects recur repeatedly, we illustrate
the implementation of just a few representative rules starting from \SComm.

@@

There are two non-trivial aspects worth commenting. The first one concerns the
proof $\AgdaInductiveConstructor{>}~q$ used in the
\AgdaInductiveConstructor{wait} process before structural pre-congruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{ch~⟨~>}~q~\AgdaInductiveConstructor{⟩}~P)$
    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
    it is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel  consumed by
    the \AgdaInductiveConstructor{wait} process (say $y$, of type $\Bot$).
    %
    We enforce the side condition $x \ne y$ of \SWait (which we left implicit in
    \Cref{tab:semantics}) imposing that the type $A$ in front of
    $A~\AgdaInductiveConstructor{∷}~\Context_1$ goes to the right partition of
    the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$ through the
    use of \AgdaInductiveConstructor{>}.
\end{enumerate}

@@

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, meaning
that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, not shown here but similar to
\AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$
is replaced by $\AgdaInductiveConstructor{>}~r$.

### 2025-12-25 paper/Context.lagda.tex
global revision

We are going to adopt a nameless representation of channels hence typing
contexts are represented using lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts as in
\Cref{sec:typing-rules}, even though in the Agda formalisation they are lists
instead of finite maps.

@@

When splitting a context $\Context$ into $\ContextD + \ContextE$, for each type
in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<}
and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
placed in $\ContextD$ or $\ContextE$.
%
For example, below is a proof of the splitting $\Splitting{[A, B, C,
D]}{[B]}{[A, C, D]}$.

\begin{code}[hide]
module _ where
  postulate A B C D : Type
\end{code}
\begin{code}
  splitting-example₁ : (A ∷ B ∷ C ∷ D ∷ []) ≃ (B ∷ []) + (A ∷ C ∷ D ∷ [])
  splitting-example₁ = > < > > •
\end{code}

@@

If $P$ and $Q$ are predicates over contexts, the predicate $P \sand Q$ holds for
those contexts $\ContextC$ that can be split into $\ContextD$ and $\ContextE$
such that $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
proofs of $P~\ContextD$ and $Q~\ContextE$.
%
The use of metavariables $P$ and $Q$ for denoting predicates over contexts is
appropriate: as we will see shortly, in our formalisation processes are indeed
an example of predicate over typing contexts.

@@

To conclude the implementation of typing contexts, we define a predicate
\AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, namely those
contexts solely made of types of the form $?A$. We will need this predicate in
the definition of a server, which must obey the typing rule \ServerRule of
\Cref{tab:typing-rules}.

>>>>

We are going to adopt a nameless representation of channels. Hence, typing
contexts are represented as lists of types, where the (polymorphic)
\AgdaDatatype{List} type and its constructors \AgdaInductiveConstructor{[]} and
\AgdaInductiveConstructor{\_∷\_} are defined in the module
\AgdaModule{Data.List} of Agda's standard library. We will keep using
$\ContextC$, $\ContextD$ and $\ContextE$ to range over typing contexts, even
though in the Agda formalisation they are lists and not finite maps as in
\Cref{sec:typing-rules}.

@@

When splitting a context $\Context$ into $\ContextD + \ContextE$, for each type
in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<}
and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
placed in $\ContextD$ or in $\ContextE$.
%
For example, below is a proof of the splitting $\Splitting{[A, B, C,
D]}{[B]}{[A, C, D]}$.

\begin{code}[hide]
module _ where
  postulate A B C D : Type
\end{code}
\begin{code}
  splitting-example₁ : (A ∷ B ∷ C ∷ D ∷ []) ≃ [ B ] + (A ∷ C ∷ D ∷ [])
  splitting-example₁ = > < > > •
\end{code}

@@

If $P$ and $Q$ are predicates over contexts, the predicate $P \sand Q$ holds for
those contexts $\ContextC$ that can be split into $\ContextD$ and $\ContextE$ so
that $P$ holds for $\ContextD$ and $Q$ holds for $\ContextE$.
%
The constructor \AgdaInductiveConstructor{\_⟨\_⟩\_} has three explicit arguments
witnessing the splitting $\Splitting\ContextC\ContextD\ContextE$ along with
proofs of $P~\ContextD$ and $Q~\ContextE$.
%
The use of metavariables $P$ and $Q$ for denoting predicates over contexts is
appropriate: as we will see shortly, in our formalisation processes are indeed
an example of predicate over typing contexts.

@@

To conclude the implementation of typing contexts, we define a predicate
\AgdaDatatype{Un} that holds for \emph{unrestricted} contexts, those solely made
of types of the form $?A$. We need this predicate in the definition of a server,
which must comply with the typing rule \ServerRule in \Cref{tab:typing-rules}.

### 2025-12-25 paper/Examples.lagda.tex
global revision

In this section we revisit and expand the sample processes discussed in
\Cref{ex:booleans,ex:echo} and show their encoding in our formalisation. The
encoding of $\Bool$ is straightforward.

@@

The function \AgdaFunction{\$} is defined in Agda's standard library and denotes
function application. We use it here as ``separator'' to save a few parentheses.
%
Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
produces a boolean value with a process that consumes a boolean value and the
use of \AgdaInductiveConstructor{link} for copying a boolean value from one
channel to another. 
%
Here we have chosen to define the disjunction \AgdaFunction{||} from the
conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
laws.

>>>>

In this section we revisit the processes discussed in \Cref{ex:booleans,ex:echo}
and show their encoding in our formalisation. The encoding of $\Bool$ is
straightforward.

@@

The function \AgdaFunction{\$} (defined in Agda's standard library) is just a
visible function application operator. We use it here as ``separator'' to save a
few parentheses.
%
Note the use of \AgdaInductiveConstructor{cut}s for combining a process that
produces a boolean value with a process that consumes a boolean value and the
use of \AgdaInductiveConstructor{link} for copying a boolean value from one
channel to another. 
%
Here we have chosen to define the disjunction \AgdaFunction{||} from the
conjunction \AgdaFunction{\&\&} and negation \AgdaFunction{!!} using De Morgan's
laws.

### 2025-12-25 paper/Permutations.lagda.tex
global revision

According to our nameless representation of channels, it is the \emph{order} of
channels in a typing context $\Context$ that determines the binding location in
the structure of a process. When the binding structure of a process changes, \eg
because a structural pre-congruence rule is applied, or when a channel
substitution occurs, \cf the right-hand side of the \LinkRule reduction in
\Cref{tab:semantics}, the typing context must be suitably rearranged. Such
rearrangement is in fact a \emph{permutation} of the elements of $\Context$.

@@

The definition of the data type \AgdaDatatype{\_↭\_} is nearly the same found in
the module \AgdaModule{Data.List.Relation.Binary.Permutation.Propositional} of
Agda's standard library. We have preferred defining our own notion of
permutation for improved simplicity and convenience: the
\AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
the tail of the typing context, which can always be performed, if needed,
combining \AgdaInductiveConstructor{swap} with \AgdaInductiveConstructor{prep}
and \AgdaInductiveConstructor{trans}. Also, and more importantly, our data type
\AgdaDatatype{\_↭\_} is monomorphic (it does not need to relate arbitrary lists)
and the arguments $A$ and $B$ of \AgdaInductiveConstructor{swap} and
\AgdaInductiveConstructor{prep} are implicit, which streamlines the usage of
these constructors in the rest of the code.

>>>>

According to our nameless representation of channels, it is the \emph{position}
of a type in a typing context $\Context$ that determines the location of its
binder in the structure of a process. When the binding structure of a process
changes, \eg because a structural pre-congruence rule is applied, or when a
channel substitution occurs, \cf the right-hand side of the \LinkRule reduction
in \Cref{tab:semantics}, the typing context must be suitably rearranged. Such
rearrangement is in fact a \emph{permutation} of the elements of $\Context$.

@@

The definition of the data type \AgdaDatatype{\_↭\_} is nearly the same found in
the module \AgdaModule{Data.List.Relation.Binary.Permutation.Propositional} of
Agda's standard library. We have preferred defining our own notion of
permutation for simplicity and convenience: the \AgdaInductiveConstructor{swap}
constructor does not need a sub-permutation for the tail of the typing context,
which can always be performed, if needed, combining
\AgdaInductiveConstructor{swap} with \AgdaInductiveConstructor{prep} and
\AgdaInductiveConstructor{trans}. Also, and more importantly, our data type
\AgdaDatatype{\_↭\_} is monomorphic (it does not need to relate arbitrary lists)
and the arguments $A$ and $B$ of \AgdaInductiveConstructor{swap} and
\AgdaInductiveConstructor{prep} are implicit, which streamlines the usage of
these constructors in the rest of the code.

### 2025-12-25 paper/Process.lagda.tex
global revision

The constructor \AgdaInductiveConstructor{link} models a process of the form
$\Link\x\y$. This process is well typed in a context of the form $x : A, y :
\dual{A}$, namely a context satisfying the predicate
$\AgdaDatatype{Ch}~A~\AgdaDatatype{∗}~\AgdaDatatype{Ch}~(\AgdaFunction{dual}~A)$
which we see on the left-hand side of \AgdaFunction{⇒}.
%
The constructor \AgdaInductiveConstructor{cut} models a cut $\Cut[A]\x{P}{Q}$.
For this process to be well typed in a context $\Context$, it must be the case
that $\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed
in the contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$ respectively.
Since $x$ is the most recently introduced channel, the types $A$ and $\dual{A}$
must be added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
means of the functions $(A~\AgdaInductiveConstructor{∷\_})$ and
$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$ (these are partial
applications of the constructor $\AgdaInductiveConstructor{\_∷\_}$ for lists to
which we have supplied the left operand).

@@

We conclude this module proving that permutations preserve process typing. Since
list permutations basically correspond to channel renaming, we can read this
property as the fact that typing is preserved by (injective) name
substitutions.

>>>>

The constructor \AgdaInductiveConstructor{link} models a process of the form
$\Link\x\y$. This process is well typed in a context of the form $x : A, y :
\dual{A}$, namely a context satisfying the predicate
$\AgdaDatatype{Ch}~A~\AgdaDatatype{∗}~\AgdaDatatype{Ch}~(\AgdaFunction{dual}~A)$
which we see on the left-hand side of \AgdaFunction{⇒}.
%
The constructor \AgdaInductiveConstructor{cut} models a cut $\Cut[A]\x{P}{Q}$.
For this process to be well typed in a context $\Context$, it must be the case
that $\Splitting\Context\ContextD\ContextE$ so that $P$ and $Q$ are well typed
in the contexts $x : A, \ContextD$ and $x : \dual{A}, \ContextE$ respectively.
Since $x$ is the most recently introduced channel, the types $A$ and $\dual{A}$
is added \emph{at the front} of $\ContextD$ and $\ContextE$, which we do by
means of the functions $(A~\AgdaInductiveConstructor{∷\_})$ and
$(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷\_})$ (these are partial
applications of the constructor $\AgdaInductiveConstructor{\_∷\_}$ for lists to
which we have supplied the left operand).

@@

We conclude this module proving that permutations preserve process typing. Since
list permutations basically correspond to channel renaming, we can read this
property as the fact that typing is preserved by (bijective) name substitutions.

### 2025-12-25 paper/Reduction.lagda.tex
global revision

Just like structural precongruence, reduction is formalised as a binary relation
between processes that are well typed in the very same typing context. Thus, the
definition of reduction embeds the subject reduction result.

@@

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
only free channel occurring in a process of the form $\Close\x$, the
\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
$(\AgdaInductiveConstructor{<}~q_0)$ to be sure that the referenced channel is
also the restricted one, and therefore matches the one of the
\AgdaInductiveConstructor{close} process.

>>>>

Just like structural precongruence, reduction is formalised as a binary relation
between processes that are well typed in the very same typing context. Thus, the
definition of reduction embeds the subject reduction result (\Cref{thm:red}).

@@

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•}$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
only free channel occurring in a process of the form $\Close\x$, the
\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
$\AgdaInductiveConstructor{<}~q_0$ to be sure that the referenced channel is
also the restricted one, and therefore matches the one of the
\AgdaInductiveConstructor{close} process.

### 2025-12-25 paper/Type.lagda.tex
global revision

The representation of types is standard, with one constructor for each of the
forms described in \Cref{sec:types}. We start by defining \emph{pre-types} as an
indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
type variables and we use de Bruijn indices for denoting type variables. In this
way, we make sure that pre-types are well scoped.

@@

Note the constructors \AgdaInductiveConstructor{var} and
\AgdaInductiveConstructor{rav}, which respectively represent type variables and
their dual, and the quantifiers \AgdaInductiveConstructor{`∀} and
\AgdaInductiveConstructor{`∃} which increase the number of bound type variables
in the scoped pre-type.

@@

This property is so important in the rest of the formalisation that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or
within a pragma in the module's source code.}

@@

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a (pre)type with other (pre)types. Even though in practice
we will always substitute one variable at a time, it is technically easier to
substitutes \emph{all} variables of a (pre)type. The definition of
\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
variables and \emph{lifting} substitutions across quantifiers. All such
definitions are straightforward adaptations of those described by
\citet{KokkeSiekWadler20}.

@@

Among all substitutions, we will use the one that substitutes the 0-indexed type
variable with a (pre)type. It is convenient to introduce this substitution once
and for all, which we do here.

@@

At last we are ready to define the type of closed pre-types, those
having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
From now on, we will seldom use pre-types again.

>>>>

The representation of types is standard, with one constructor for each of the
forms described in \Cref{sec:types}. We start by defining an indexed data type
$\AgdaDatatype{PreType}~n$ to represents (\LCP) types in the scope of $n$
quantifiers and we use elements of $\AgdaDatatype{Fin}~n$ as de Bruijn indices
for the quantified type variables. In this way, we make sure that pre-types are
well scoped.

@@

Note the constructors \AgdaInductiveConstructor{var} and
\AgdaInductiveConstructor{rav}, which respectively represent type variables and
their dual, and the quantifiers \AgdaInductiveConstructor{`∀} and
\AgdaInductiveConstructor{`∃} which increase the number of quantifiers in the
scoped pre-type.

@@

This property is important in the rest of the formalisation so we define an
implicit rewriting rule that Agda can autonomously apply whenever possible. This
is achieved by means of the following pragma directive.\footnote{The directive
is effective provided that the option \texttt{{-}{-}rewriting} is enabled,
either globally when invoking Agda or within a pragma in the module's source
code.}

@@

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a pre-type with other pre-types. In practice we will
always substitute one variable at a time, but it is technically easier to
substitute \emph{all} variables of a pre-type, possibly with themselves. The
definition of \AgdaFunction{subst} relies on some auxiliary functions for
\emph{renaming} type variables and \emph{lifting} substitutions across
quantifiers. These functions are straightforward adaptations of those described
by \citet{KokkeSiekWadler20}.

@@

Among all substitutions, we will use the one that substitutes the 0-indexed type
variable with a pre-type. It is convenient to introduce this substitution once
and for all, which we do here.

@@

We call \AgdaDatatype{Type} closed pre-types, those having no free type
variables. From now on, we will seldom use pre-types again.

### 2025-12-25 paper/calculus.tex
global revision

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalising two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) the peer endpoints of the same session are always used in
complementary ways. This latter property is called \emph{well formedness} in the
challenge. In this work we also consider \emph{deadlock freedom}, which is more
general than well formedness and holds for \Calculus since its type system is
based on linear logic. We now formulate these properties using the notation
developed so far.

>>>>

The linearity challenge~\citep{CarboneEtAl24} aims at formalising two essential
properties of well-typed processes: (1) typing is preserved by reductions; (2)
the peer endpoints of the same session are always used in complementary ways.
This latter property is called \emph{well formedness} in the challenge. In this
work we also consider \emph{deadlock freedom}, which is more general than well
formedness and holds for \Calculus since its type system is based on linear
logic. We now formulate these properties using the notation developed so far.

### 2025-12-25 paper/conclusion.tex
global revision

We have presented a formalisation of \LCP, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCP using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalisation and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite interesting to note that, among the approaches that have been
proposed to overcome context splitting, the one of \citet{SanoKavanaghPientka23}
makes use of explicit continuations.

@@

The simple formalisation of \LCP is a good basis for further investigations. We
have already extended \LCP with support for infinite types and recursive
processes (the source code of this extension is publicly available in a sibling
folder of the one for \LCP). In the future, it would be interesting to formalise
the strong normalisation property of \LCP as a consequence of cut elimination of
classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those of \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which are significantly more complex than those
of binary sessions. Just to give an example, the formalisation by
\citet{TiroreBengstonCarbone25} amounts to more than 1Mb of Coq code.
Incidentally, multiparty sessions are not supported by a strong connection with
linear logic and there is no known continuation-passing encoding of multiparty
sessions. Hence, it is unclear whether the considerations made for this work
also apply to the multiparty case.

>>>>

We have presented a formalisation of \LCP, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCP using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCP can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalisation and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite striking that, among the approaches that have been proposed to
overcome context splitting, the one of \citet{SanoKavanaghPientka23} makes use
of explicit continuations.

@@

The simple formalisation of \LCP is a good basis for further investigations. We
have already extended \LCP with support for infinite types and recursive
processes (the source code of this extension is publicly available in the same
repository as the one that contains the formalisation of \LCP). In the future,
it would be interesting to formalise the strong normalisation property of \LCP
as a consequence of cut elimination of classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which can be significantly more complex than
those of binary sessions. The formalisation by \citet{JacobsBalzerKrebbers22a}
amounts to 173kb and the one by \citet{TiroreBengstonCarbone25} to more than 1Mb
of Coq code. Also in these cases, the formalisation based on (virtual)
continuations~\citep{JacobsBalzerKrebbers22a} happens to be substantially
smaller. Whether this is a coincidence or further evidence of the effectiveness
of the continuation-based approaches is left for future work.
